---
title: "ranked-gene-stability"
author: "S2599932"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Overview: Gene Stability and Enrichment Analysis in Single-Cell RNA Sequencing Data

This project aims to compare gene stability and enrichment between different single-cell RNA sequencing (scRNA-seq) datasets. Specifically, we will:
1. Identify the top 1000 stable genes in two scRNA-seq datasets.
2. Compare the rank changes of these genes between the datasets.
3. Use rank changes as input for enrichment analysis using the `fgsea` package.
4. Validate the methodology by fitting models to different embryonic stem cell (ES) datasets and comparing enrichment results.

### Data Sources

- **ES Study Data (E-MTAB-2600):** Used in the Functional Genomics and Transcriptomics (FGT) class.
- **Early Development Data (GSE121708):** Counts table from stages E4.5, E5.5, E6.5, E7.5, sequenced with scNMT-seq from Argelaguet et al., Nature 2019.

### Methodology

1. **Fit the scSEGIndex Model:**
   - Apply the scSEGIndex model to the ES dataset (E-MTAB-2600) and the early development dataset (GSE121708) to identify the top 1000 stable genes.

2. **Rank Change Comparison:**
   - Compare the ranks of the top 1000 stable genes between the two datasets.
   - Rank the genes by the amount and direction of rank change.

3. **Enrichment Analysis:**
   - Use the ranked list of genes as input for `fgsea` (fast gene set enrichment analysis).
   - Compare the results against the Molecular Signatures Database (MSigDB).
   - Perform permutation tests to search for enrichment in the sorted list.

4. **Validation with Different ES Cell Types:**
   - Apply the scSEGIndex model to two different types of ES cells.
   - Repeat the enrichment comparison to validate the methodology.

### Data Files

- **E-MTAB-2600 Dataset:** Available on `bioinfmsc6` in the `/shared_files/FGT_T9` folder.
- **GSE121708 Dataset:** Counts data from GSE121708 record.

### Next Steps

1. **Data Preprocessing:**
   - Load and preprocess the scRNA-seq data from the specified sources.

2. **Model Fitting:**
   - Fit the scSEGIndex model to both datasets to identify the top 1000 stable genes.

3. **Rank Change Analysis:**
   - Compute and compare the rank changes of the top 1000 stable genes between the datasets.

4. **Enrichment Analysis:**
   - Perform `fgsea` on the ranked list of genes and compare results to MSigDB.
   - Conduct permutation tests to assess enrichment.

5. **Validation:**
   - Apply the methodology to different ES cell types and compare enrichment results.

6. **Results Interpretation:**
   - Analyze and interpret the results of the enrichment analysis.
   - Prepare the data for further analysis using David.

This project demonstrates a proof-of-concept for comparing gene stability and enrichment between scRNA-seq datasets. The methodology will be further refined and organized for production use.



```{r load_libraries, message=FALSE,warning=FALSE}
library(SingleCellExperiment)
library(scMerge)
library(ExperimentHub)
library(GSEABase)
library(fgsea)
library(msigdb)

library(Seurat)
library(dplyr)
```

### Create **SingleCellExperiment** objects and filter them based on QC metrics

```{r create_early_development_SCE, message=FALSE,warning=FALSE}
# The counts matrix was created from the E4.5, E5.5, E6.5, E7.5 stages of mouse devlopment
dev_counts <- read.table("../dissertation-prolouge/TestData/counts.txt",sep="\t",header=T,row.names=1)
dev_targets <- read.table("../dissertation-prolouge/TestData/sample_metadata.txt", header=1, sep = "\t")
dev_counts_ordered <- dev_counts[,dev_targets$id_rna]

# Create the SCE object
dev_sce_raw <- SingleCellExperiment(assays = list(counts = as.matrix(dev_counts_ordered)))
# Add metadata to dev_sce_raw
colData(dev_sce_raw) <- DataFrame(dev_targets[match(colnames(dev_sce_raw), dev_targets$id_rna), ])
```



```{r create_mouse_ESC_SCE, message=FALSE, warning=FALSE}
# Read in the RDS file for class data
es_raw <- readRDS("../dissertation-prolouge/Data/Class_Data/single_es_tximport.rds")
es_targets <- read.table("../dissertation-prolouge/Data/Class_Data/E-MTAB-2600.targets.txt", sep = "\t", header = TRUE, row.names = "ERR")

# Create the ESC SCE object
es_sce <- SingleCellExperiment(assays=list(counts = es_raw$counts, abundance = es_raw$abundance))
colData(es_sce) <- DataFrame(es_targets[match(colnames(es_sce), rownames(es_targets)), ]) # Add metadata


# Calculate QC metrics
is_spike <- grepl("^ERCC",rownames(es_sce))
mitogenes <- "ENSMUSG00000064336|ENSMUSG00000064337|ENSMUSG00000064338|ENSMUSG00000064339|ENSMUSG00000064340|ENSMUSG00000064341|ENSMUSG00000064342|ENSMUSG00000064343|ENSMUSG00000064344|ENSMUSG00000064345|ENSMUSG00000064346|ENSMUSG00000064347|ENSMUSG00000064348|ENSMUSG00000064349|ENSMUSG00000064350|ENSMUSG00000064351|ENSMUSG00000064352|ENSMUSG00000064353|ENSMUSG00000064354|ENSMUSG00000064355|ENSMUSG00000064356|ENSMUSG00000064357|ENSMUSG00000064358|ENSMUSG00000064359|ENSMUSG00000064360|ENSMUSG00000064361|ENSMUSG00000065947|ENSMUSG00000064363|ENSMUSG00000064364|ENSMUSG00000064365|ENSMUSG00000064366|ENSMUSG00000064367|ENSMUSG00000064368|ENSMUSG00000064369|ENSMUSG00000064370|ENSMUSG00000064371|ENSMUSG00000064372"
is_mito <- grepl(mitogenes, rownames(es_sce))
es_sce$qc_stats <- perCellQCMetrics(es_sce, subsets=list(ERCC=is_spike, Mt=is_mito))

# Calculating the filtering metrics
libsize_drop <- isOutlier(es_sce$qc_stats$sum, nmads = 3, type = "lower", log = TRUE)
feature_drop <- isOutlier(es_sce$qc_stats$detected, nmads = 3, type = "lower", log = TRUE)
mito_drop <- isOutlier(es_sce$qc_stats$subsets_Mt_percent, nmads = 3, type = "higher")
spike_drop <- isOutlier(es_sce$qc_stats$subsets_ERCC_percent, nmads = 3, type = "higher")
 
es_sce_copy <- es_sce # Save a non filtered copy

# Filtering the SingleCellExperiment(SCE) object
es_sce_fil <- es_sce[,!(libsize_drop|feature_drop|mito_drop|spike_drop)]
es_sce_fil_copy <- es_sce_fil # Save a filtered copy

# Save the filtered cells
filtered_cells <- colnames(es_sce_fil)
```

> **es_sce_fil** should be used for downstream analysis and **es_sce_fil_copy** has been created to keep a copy for other uses.

```{r deconvolution_normalization_of_development_data, message=FALSE, warning=FALSE}

```





















































































































