---
title: "pca-finalized"
author: "S2599932"
date: "2024-07-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(scran)
library(scater)
library(biomaRt)
library(scMerge)
library(cowplot)
```

```{r}
data_dir <- "/home/s2599932/study-datasets/mouse-dataset/E-MTAB-2600/single_es_tximport.rds"
annotation_dir <- "/home/s2599932/study-datasets/mouse-dataset/E-MTAB-2600/E-MTAB-2600.targets.txt"
```

```{r}
raw <- readRDS(data_dir)
annotation <- read.table(annotation_dir, sep = "\t", header = TRUE, row.names = "ERR")
annotation <- annotation[match(colnames(raw$counts), rownames(annotation)), ]
```

```{r}
data("segList_ensemblGeneID", package = "scMerge")
mSEG <- segList_ensemblGeneID$mouse$mouse_scSEG
```

```{r}
# Create Single Cell Experiment object
sce <- SingleCellExperiment(
  assays = list(counts = as.matrix(raw$counts)),
  colData = annotation
)
```

## Quality control
```{r}
is_spike <- grepl("^ERCC", rownames(sce))
mitogenes <- "ENSMUSG00000064336|ENSMUSG00000064337|ENSMUSG00000064338|ENSMUSG00000064339|ENSMUSG00000064340|ENSMUSG00000064341|ENSMUSG00000064342|ENSMUSG00000064343|ENSMUSG00000064344|ENSMUSG00000064345|ENSMUSG00000064346|ENSMUSG00000064347|ENSMUSG00000064348|ENSMUSG00000064349|ENSMUSG00000064350|ENSMUSG00000064351|ENSMUSG00000064352|ENSMUSG00000064353|ENSMUSG00000064354|ENSMUSG00000064355|ENSMUSG00000064356|ENSMUSG00000064357|ENSMUSG00000064358|ENSMUSG00000064359|ENSMUSG00000064360|ENSMUSG00000064361|ENSMUSG00000065947|ENSMUSG00000064363|ENSMUSG00000064364|ENSMUSG00000064365|ENSMUSG00000064366|ENSMUSG00000064367|ENSMUSG00000064368|ENSMUSG00000064369|ENSMUSG00000064370|ENSMUSG00000064371|ENSMUSG00000064372"
is_mito <- grepl(mitogenes, rownames(sce))
sce$qc_stats <- perCellQCMetrics(sce, subsets=list(ERCC=is_spike, Mt=is_mito))

# Calculate the QC metrics based on which cells will be retained or eliminated
libsize_drop <- isOutlier(sce$qc_stats$sum, nmads = 3, type = "lower", log = TRUE)
feature_drop <- isOutlier(sce$qc_stats$detected, nmads = 3, type = "lower", log = TRUE)
mito_drop <- isOutlier(sce$qc_stats$subsets_Mt_percent, nmads = 3, type = "higher")
spike_drop <- isOutlier(sce$qc_stats$subsets_ERCC_percent, nmads = 3, type = "higher")

# Save a copy of the unfiltered SCE object
sce_copy <- sce

# Drop the samples from the SCE object that do not meet the QC standards
sce_fil <- sce[, !(libsize_drop | feature_drop | mito_drop | spike_drop)]
filtered_cells <- colnames(sce_fil)
```

## Normalize using deconvolution normalization
```{r}
clusters <- quickCluster(sce_fil)
sce_fil <- computeSumFactors(sce_fil, clusters = clusters)
sce_fil <- logNormCounts(sce_fil)
```

## Exclude the outlying cells (only can be done once the outliers has been detected)
```{r}
# Read in the outliers
outliers_df <- read.csv("./emtab2600-pca-outliers.csv", header = TRUE)
outliers_list <- outliers_df$X

# Remove the outlying cells from the sce_object
sce_fil <- sce_fil[, !colnames(sce_fil) %in% outliers_list]
```

# PCA using Seurat's RunPCA() function
## Create seurat object
```{r}
log_counts <- assay(sce_fil, "logcounts")
  
# Creating the Seurat object(so)
so <- CreateSeuratObject(counts = log_counts)
so <- AddMetaData(so, metadata = annotation)

DefaultAssay(so) <- "RNA"
so <- SetAssayData(so, layer = "data", new.data = log_counts)
```

```{r}
so_fil <- so[,filtered_cells]
so_fil <- ScaleData(so_fil, verbose = FALSE, features = rownames(so_fil))
so_fil <- FindVariableFeatures(so_fil, selection.method = "vst", verbose = FALSE)
```

```{r}
# Only include single cells
# Convert the metadata to a tibble for filtering
metadata <- so_fil@meta.data %>%
  rownames_to_column(var = "cell")

# Filter the metadata to only include the desired cell types
filtered_metadata <- metadata %>%
  filter(Type %in% c("2i", "a2i", "serum"))

# Subset the Seurat object using the filtered metadata
so_fil <- subset(so_fil, cells = filtered_metadata$cell)
```

```{r}
# Make copy of the Seurat object for doing PCA with different combination of parameters
so_non_subsetted_no_features_defined <- so_fil
so_non_subsetted_all_genes_as_features <- so_fil
so_non_subsetted_mSEGs_as_features <- so_fil
so_subsetted_no_features_defined <- so_fil
so_subsetted_mSEGs_as_features <- so_fil
```

## Run only once (mainly to identify the outliers)
```{r fig.height=8, fig.width=10}
so_non_subsetted_no_features_defined <- RunPCA(so_non_subsetted_no_features_defined, verbose = FALSE)

# Extract PCA embeddings and metadata, and combine them
pca_df <- Embeddings(so_non_subsetted_no_features_defined, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_non_subsetted_no_features_defined@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_non_subsetted_no_features_defined, reduction = "pca")^2 / sum(Stdev(so_non_subsetted_no_features_defined, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
pca_plot <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right",
    legend.key.size = unit(0.8, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "PCA of E-MTAB-3929 using hSEGs as features",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )


print(pca_plot)
```

## Identify the outlying cells and remove them from the PCA plots
```{r}
non_subsetted_no_features_defined_embeddings <- Embeddings(so_non_subsetted_no_features_defined, "pca")[,1:2]
pca_embedding <- as.data.frame(non_subsetted_no_features_defined_embeddings)

outliers <- pca_embedding %>% arrange(PC_1) %>% head(7)
outliers_list <- rownames(outliers)

write.csv(outliers, file = "/home/s2599932/MSc_DSB_Dissertation_2024/01-reproduced-old-dissertation/finalized-pca-workflow/emtab2600-pca-outliers.csv")
```

## Read in the outliers
```{r}
outliers_df <- read.csv("/home/s2599932/MSc_DSB_Dissertation_2024/01-reproduced-old-dissertation/finalized-pca-workflow/emtab2600-pca-outliers.csv", header = TRUE)
outliers_list <- outliers_df$X
```

## Plots
### Plot 1
```{r}
# Remove the outlying cells
so_non_subsetted_no_features_defined <- subset(so_non_subsetted_no_features_defined, cells = setdiff(Cells(so_non_subsetted_no_features_defined), outliers_list))

so_non_subsetted_no_features_defined <- RunPCA(so_non_subsetted_no_features_defined, verbose = FALSE)

# Extract PCA embeddings and metadata, and combine them
pca_df <- Embeddings(so_non_subsetted_no_features_defined, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_non_subsetted_no_features_defined@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_non_subsetted_no_features_defined, reduction = "pca")^2 / sum(Stdev(so_non_subsetted_no_features_defined, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
p1 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

ggsave(filename = "./seurat_p1.png", plot = p1, width = 10, height = 8, dpi = 300)

print(p1)
```

### Plot 2
```{r}
# Remove the outlying cells
so_non_subsetted_all_genes_as_features <- subset(so_non_subsetted_all_genes_as_features, cells = setdiff(Cells(so_non_subsetted_all_genes_as_features), outliers_list))

so_non_subsetted_all_genes_as_features <- RunPCA(so_non_subsetted_all_genes_as_features, features = rownames(so_non_subsetted_all_genes_as_features), verbose = FALSE)

# Extract PCA embeddings and metadata, and combine them
pca_df <- Embeddings(so_non_subsetted_all_genes_as_features, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_non_subsetted_all_genes_as_features@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_non_subsetted_all_genes_as_features, reduction = "pca")^2 / sum(Stdev(so_non_subsetted_all_genes_as_features, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
p2 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

ggsave(filename = "./seurat_p2.png", plot = p2, width = 10, height = 8, dpi = 300)

print(p2)
```

### Combine plot 1 and 2
```{r}
# Load the necessary libraries
library(cowplot)

# Combine the plots into one object using plot_grid and add labels
combined_plot <- plot_grid(p1, p2, labels = c("A", "B"), ncol = 2)

# Save the combined plot as a high-resolution PNG
ggsave(filename = "seurat-p1vsp2.png", plot = combined_plot, width = 16, height = 8, dpi = 300)
```

### Plot 3
```{r}
# Remove the outlying cells
so_non_subsetted_mSEGs_as_features <- subset(so_non_subsetted_mSEGs_as_features, cells = setdiff(Cells(so_non_subsetted_mSEGs_as_features), outliers_list))

so_non_subsetted_mSEGs_as_features <- RunPCA(so_non_subsetted_mSEGs_as_features, features = mSEG, verbose = FALSE)

# Extract PCA embedding and metadata, and combine them
pca_df <- Embeddings(so_non_subsetted_mSEGs_as_features, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_non_subsetted_mSEGs_as_features@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_non_subsetted_mSEGs_as_features, reduction = "pca")^2 / sum(Stdev(so_non_subsetted_mSEGs_as_features, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
p3 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "PCA of E-MTAB-2600 mSEGs as features",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

ggsave(filename = "./seurat_p3.png", plot = p3, width = 10, height = 8, dpi = 300)

print(p3)
```

### Plot 4
```{r}
# Remove the outlying cells
so_subsetted_no_features_defined <- subset(so_subsetted_no_features_defined, cells = setdiff(Cells(so_subsetted_no_features_defined), outliers_list))
# Subset the Seurat object to only contain mSEG rows
so_subsetted_no_features_defined <- subset(so_subsetted_no_features_defined, features = mSEG)
so_subsetted_no_features_defined <- RunPCA(so_subsetted_no_features_defined, verbose = FALSE)

# Extract PCA embeddings and metadata, and combine them
pca_df <- Embeddings(so_subsetted_no_features_defined, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_subsetted_no_features_defined@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_subsetted_no_features_defined, reduction = "pca")^2 / sum(Stdev(so_subsetted_no_features_defined, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
p4 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

ggsave(filename = "./seurat_p4.png", plot = p4, width = 10, height = 8, dpi = 300)

print(p4)

```


### Plot 5
```{r}
# Remove the outlying cells
so_subsetted_mSEGs_as_features <- subset(so_subsetted_mSEGs_as_features, cells = setdiff(Cells(so_subsetted_mSEGs_as_features), outliers_list))
# Subset the Seurat object to only contain mSEG rows
so_subsetted_mSEGs_as_features <- subset(so_subsetted_mSEGs_as_features, features = mSEG)
so_subsetted_mSEGs_as_features <- RunPCA(so_subsetted_mSEGs_as_features, features = mSEG, verbose = FALSE)

# Extract PCA embedding and metadata, and combine them
pca_df <- Embeddings(so_subsetted_mSEGs_as_features, reduction = "pca")[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "cell") %>%
  left_join(so_subsetted_mSEGs_as_features@meta.data %>%
              rownames_to_column(var = "cell"),
            by = "cell")

# Calculate the variance explained by the first two principal components (for labels)
explained_variance <- Stdev(so_subsetted_mSEGs_as_features, reduction = "pca")^2 / sum(Stdev(so_subsetted_mSEGs_as_features, reduction = "pca")^2) * 100
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2 using the specified parameters
p5 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

ggsave(filename = "./seurat_p5.png", plot = p5, width = 10, height = 8, dpi = 300)

print(p5)
```

### Combine plot 4 and 5
```{r}
# Combine the plots into one object using plot_grid and add labels
combined_plot <- plot_grid(p4, p5, labels = c("C", "D"), ncol = 2)

# Save the combined plot as a high-resolution PNG
ggsave(filename = "seurat-p4vsp5.png", plot = combined_plot, width = 16, height = 8, dpi = 300)
```

# Next create heatmap of the log normalized data being used in the PCA for subsetted seurat object with and without explicit defination of mSEG as features
```{r}

```


---

# PCA using runPCA() of scater
```{r}
# Remove the bulk samples and only include single cell samples
metadata <- as.data.frame(colData(sce_fil)) %>%
  rownames_to_column(var = "cell")

filtered_metadata <- metadata %>%
  filter(Type %in% c("2i", "a2i", "serum"))

sce_fil <- sce_fil[, colnames(sce_fil) %in% filtered_metadata$cell]

# Copy the SCE objects
sce_non_subsetted_no_features_defined <- sce_fil
sce_non_subsetted_all_genes_as_features <- sce_fil
sce_non_subsetted_mSEGs_as_features <- sce_fil
sce_subsetted_no_features_defined <- sce_fil
sce_subsetted_mSEGs_as_features <- sce_fil
```

## Plot 1
```{r}
sce_non_subsetted_no_features_defined <- runPCA(sce_non_subsetted_no_features_defined, exprs_values = "logcounts", subset_row = NULL, scale = TRUE)

# Extract PCA embedding
pca_embeddings <- reducedDim(sce_non_subsetted_no_features_defined, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_non_subsetted_no_features_defined))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_non_subsetted_no_features_defined, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p1 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p1.png", plot = p1, width = 10, height = 8, dpi = 300)

# Print the plot
print(p1)
```






























## Plot 2
```{r}
# Perform PCA on the SCE object
sce_non_subsetted_all_genes_as_features <- runPCA(sce_non_subsetted_all_genes_as_features, exprs_values = "logcounts", subset_row = rownames(sce_non_subsetted_all_genes_as_features), scale = TRUE)

# Extract PCA embeddings
pca_embeddings <- reducedDim(sce_non_subsetted_all_genes_as_features, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_non_subsetted_all_genes_as_features))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_non_subsetted_all_genes_as_features, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p2 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p2.png", plot = p2, width = 10, height = 8, dpi = 300)

# Print the plot
print(p2)

```
## Combine Plot 1 and 2
```{r}
combined_plot <- plot_grid(p1, p2, labels = c("A", "B"), ncol = 2)
ggsave(filename = "scater-p1vsp2.png", plot = combined_plot, width = 16, height = 8, dpi = 300)
```
## Plot 3
```{r}
# Create valid row indices
mSEG_index <- intersect(mSEG, rownames(sce_non_subsetted_mSEGs_as_features))

# Perform PCA on the SCE object
sce_non_subsetted_mSEGs_as_features <- runPCA(sce_non_subsetted_mSEGs_as_features, exprs_values = "logcounts", subset_row = mSEG_index, scale = TRUE)

# Extract PCA embeddings
pca_embeddings <- reducedDim(sce_non_subsetted_mSEGs_as_features, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_non_subsetted_mSEGs_as_features))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_non_subsetted_mSEGs_as_features, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p3 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p3_mSEGs.png", plot = p3, width = 10, height = 8, dpi = 300)

# Print the plot
print(p3)
```

## Plot 4
```{r}
# Create valid row indices
mSEG_index <- intersect(mSEG, rownames(sce_non_subsetted_mSEGs_as_features))

# Subset the sce object
sce_subsetted_no_features_defined <- sce_subsetted_no_features_defined[mSEG_index,]

# Perform PCA on the SCE object
sce_subsetted_no_features_defined <- runPCA(sce_subsetted_no_features_defined, exprs_values = "logcounts", subset_row = NULL, scale = TRUE)

# Extract PCA embeddings
pca_embeddings <- reducedDim(sce_subsetted_no_features_defined, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_subsetted_no_features_defined))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_subsetted_no_features_defined, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p4 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p4_no_features.png", plot = p4, width = 10, height = 8, dpi = 300)

# Print the plot
print(p4)
```

## Plot 5
```{r}
# Create valid row indices
mSEG_index <- intersect(mSEG, rownames(sce_subsetted_mSEGs_as_features))

# Subset the sce object
sce_subsetted_mSEGs_as_features <- sce_subsetted_mSEGs_as_features[mSEG_index,]

# Perform PCA on the SCE object
sce_subsetted_mSEGs_as_features <- runPCA(sce_subsetted_mSEGs_as_features, exprs_values = "logcounts", subset_row = mSEG_index, scale = TRUE)

# Extract PCA embeddings
pca_embeddings <- reducedDim(sce_subsetted_mSEGs_as_features, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_subsetted_mSEGs_as_features))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_subsetted_mSEGs_as_features, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p5 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p5_mSEGs.png", plot = p5, width = 10, height = 8, dpi = 300)

# Print the plot
print(p5)
```

## Combine Plot 4 and 5
```{r}
combined_plot <- plot_grid(p4, p5, labels = c("C", "D"), ncol = 2)
ggsave(filename = "scater-p4vsp5.png", plot = combined_plot, width = 16, height = 8, dpi = 300)
```


# PCA using fixedPCA() of scater
```{r}
# Remove the bulk samples and only include single cell samples
metadata <- as.data.frame(colData(sce_fil)) %>%
  rownames_to_column(var = "cell")

filtered_metadata <- metadata %>%
  filter(Type %in% c("2i", "a2i", "serum"))

sce_fil <- sce_fil[, colnames(sce_fil) %in% filtered_metadata$cell]

# Copy the SCE objects
sce_non_subsetted_no_features_defined <- sce_fil
sce_non_subsetted_all_genes_as_features <- sce_fil
sce_non_subsetted_mSEGs_as_features <- sce_fil
sce_subsetted_no_features_defined <- sce_fil
sce_subsetted_mSEGs_as_features <- sce_fil
```

# Plot 1
```{r}
# Perform PCA on the SCE object using fixedPCA
sce_non_subsetted_no_features_defined <- fixedPCA(sce_non_subsetted_no_features_defined, subset.row = NULL, assay.type = "logcounts", scale = TRUE)

# Extract PCA embeddings
pca_embeddings <- reducedDim(sce_non_subsetted_no_features_defined, "PCA")[, 1:2]
pca_df <- as.data.frame(pca_embeddings)
colnames(pca_df) <- c("PC_1", "PC_2")
pca_df <- rownames_to_column(pca_df, var = "cell")

# Combine with metadata
metadata <- as.data.frame(colData(sce_non_subsetted_no_features_defined))
metadata <- rownames_to_column(metadata, var = "cell")
pca_df <- left_join(pca_df, metadata, by = "cell")

# Calculate the variance explained by the first two principal components
explained_variance <- attr(reducedDim(sce_non_subsetted_no_features_defined, "PCA"), "percentVar")
pc1 <- round(explained_variance[1], 2)
pc2 <- round(explained_variance[2], 2)

# Plot with ggplot2
p1 <- ggplot(pca_df, aes(x = PC_1, y = PC_2, color = Type)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_brewer(palette = "Set2") +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold", hjust = 0.5, size = 8),
    legend.position = "right",
    legend.key.size = unit(0.6, "cm"),
    legend.text = element_text(size = 10),
    plot.margin = margin(20, 20, 20, 20)  # Add margins (top, right, bottom, left)
  ) +
  labs(
    title = "",
    x = paste("Principal Component 1 (", pc1, "%)", sep = ""),
    y = paste("Principal Component 2 (", pc2, "%)", sep = ""),
    color = "Cell Type"
  )

# Save the plot
ggsave(filename = "./scater_p1.png", plot = p1, width = 10, height = 8, dpi = 300)

# Print the plot
print(p1)
```

